---
title: "Getting baseline"
output:
  html_document:
    toc: yes
    toc_float: yes
---
```{r}
library(AIScanR)
library(spocc)
library(robis)
library(rgbif)
library(sf)
library(rnaturalearth)
library(mregions)
library(rcanvec)
library(tidyverse)
```

# Setting up basic geography

This project uses both a latitude/longitude coordinate reference system and [Statistics Canada Lambert](https://spatialreference.org/ref/epsg/nad83-statistics-canada-lambert/) projection. Let's set this up below for ease of use.

```{r}
latlong <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
#statscan
proj <- "+proj=lcc +lat_1=49 +lat_2=77 +lat_0=63.390675 +lon_0=-91.86666666666666 +x_0=6200000 +y_0=3000000 +ellps=GRS80 +datum=NAD83 +units=m +no_defs "
```

## EEZ

The EEZ is from [Marineregions.org](http://www.marineregions.org/) acquired via the `mregions` package. The code below only needs to be run if you do not yet have a locally saved version.

```{r, eval=FALSE}
EEZ <- mr_shp(key="MarineRegions:eez",maxFeatures = 12) %>%
  st_as_sf() %>%
  dplyr::filter(geoname=="Canadian Exclusive Economic Zone")
st_write(EEZ,"data/EEZ.shp")
```

Instead of the above, we load the local copy.

```{r}
EEZ <- st_read("data/EEZ.shp") %>% 
  st_combine() %>% 
  st_transform(proj)
```


## Landmass

The Canadian landmass is from [Natural Earth](https://www.naturalearthdata.com/) acquired via the `rnaturalearth` package

```{r}
Canada <- ne_states(country = c("Canada"),
                    returnclass = "sf") %>% 
  st_combine() %>% 
  st_transform(proj)
```

## Sampling grid

Since sampling for Canada as a whole in one shot would actually break either your computer or the internet, we will divide up the process into a spatial grid and do it cell by cell.

```{r, eval=FALSE}
grid <- st_make_grid(st_union(st_as_sfc(st_bbox(EEZ)),
                      st_as_sfc(st_bbox(Canada))),n = 150) %>% 
  st_as_sf() %>% 
  mutate(canada=lengths(st_intersects(.,Canada))>0,
         eez=lengths(st_intersects(.,EEZ))>0) %>% 
  filter(eez|canada)
st_write(grid,"data/grid.shp")
```

But again to save time, instead of running the above, let's just load the file it generates

```{r}
grid <- st_read("data/grid.shp")

```


## Geography Map

Let's see what we've done so far and identify a 'test' grid cell (red) to demonstrate the effectiveness of the functions below.

```{r, eval=FALSE}
# cell <- grid[193,]
g <- 339
# cell <- grid[g,]

ggplot(grid)+
  geom_sf(data=Canada,fill='lightgreen')+
  geom_sf(data=EEZ,fill='lightblue')+
  geom_sf(fill='transparent')+
  geom_sf(data=grid[g,],fill='red')
```


# Querying and cleaning

This will be a giant process in a loop, so to make it readable, tractable, and fixable, lets make some functions to break up the process. 

## iswet

The `iswet` will create a polygon delineating which parts of a grid cell are 'wet' (i.e. in the ocean, lake, wetland, etc), and provides a 1m buffer around all such features

```{r iswet, eval=FALSE}
# iswet
wet <- iswet(grid,g,EEZ,latlong,proj)


ggplot(grid[g,])+
  geom_sf(fill='transparent',colour='yellow')+
  geom_sf(data=wet,fill='lightblue',colour='lightblue')

```



## getdata

The `getdata` function will query the Ocean Biogeographic Information System ([OBIS](https://obis.org/)), the Global Biodiversity Information Facility ([GBIF](https://www.gbif.org/)), and [iNaturalist](https://inaturalist.ca)


```{r getdata, eval=FALSE}
occ <- getdata(grid, g,latlong)

head(occ)
```


## isaquatic

```{r, eval=FALSE}
test <- occ %>% 
  left_join(isaquatic(occ,wet),by="scientificName")


ggplot(grid[g,])+
  geom_sf(fill='transparent',colour='yellow')+
  geom_sf(data=wet,fill='lightblue',colour='transparent')+
  geom_sf(data=test, aes(fill=aquatic,colour=aquatic))

```



# Loop: query, clean, repeat

temporarily restrict to NS
```{r}
NS <- ne_states(country = c("Canada"),
                    returnclass = "sf") %>% 
  filter(name_en=="Nova Scotia") %>% 
  st_transform(proj) %>% 
  st_buffer(1000000)

ns <- (lengths(st_intersects(grid,NS))>0) %>% 
  which()
```



```{r}
while(TRUE){
  try(# for(g in 1:nrow(grid)){
    for(g in rev(ns)){
      print(paste("Querying  for grid:",g))
      if(sum(match(paste0("occ_",sprintf("%05d",g),"_",format(seq(Sys.time()-7*24*3600,Sys.time(),24*3600),"%Y_%m_%d"),".rds"),list.files("data/occurences/")),na.rm=TRUE)==0){
        wet <- iswet(grid,g,EEZ,latlong,proj)
        occ <- getdata(grid, g,latlong)
        isaquatic(occ,wet)
        saveRDS(occ,paste0("data/occurences/occ_",sprintf("%05d",g),"_",format(Sys.time(), "%Y_%m_%d"),".rds"))
        gc()

      }
      
    })
}

```































